Discrete Quantum Theories and Computing

### 2018-02-07 prop_prop_letter_ajhs_referee_response.tex

We understand that the argument against $\delta$-determinism by the
referee assumes that the parameter $\delta$ is \emph{exclusively}
related to the density of points (states) defined in a general Hilbert
space. This is indeed Meyer's great insight and main assumption in
his proof about the field of rationals nullifying the KS theorem.
In other words, Meyer attributes finite-precision errors to the description
of the states. We, however, make no particular assumptions about the
description of the states, and attribute finite-precision errors to
the measurement process itself as realized in a laboratory. In our
framework, the parameter $\delta$ reflects insufficient knowledge
of the experimenter which could be due to a variety of reasons related
to imperfections of devices. A typical question that an experimenter
may not be able to answer accurately would be ``did an electron land
to the left or to the right of the screen''? There are cases in which
the electron would land too close to the middle of the screen for
the experimenter to be able to determine with certainty if that is
closer to the left or the right. That imprecision might be modeled
by limiting the description of the quantum states to rational points
but we do not make this assumption. We simply record this imprecision
in the probability we assign to events and use the axioms of probability
theory (additivity, etc) to argue that a sharp transition occurs when
the imprecisions reach a certain value.

The following argument shows a concrete but simplified situation which
explains the physical significance of the parameter $\delta$. We
then follow this explanation by providing experimental results for
a three-dimensional quantum system which cannot be described by any
$\delta$-deterministic QIVPM for $\delta<\frac{1}{3}$ but can be
described by a $\frac{1}{3}$-deterministic QIVPM.

The meaning of $\delta$: To determine the probability of any event,
we typically repeat an experiment $n$ times and count the number
of times we witness the event. This assumes that for each run of the
experiment we can determine, using our apparatuses, whether the event
occurred or not. Assume an event has an ideal mathematical probability
of $0$, and we repeat the experiment $100$ times. In a perfect world
we should be able to refute the event $100$ times and calculate that
the probability is $0$. We might also observe the event $2$ times
and refute it $98$ times and therefore calculate the probability
to be $0.02$. This situation still assumes perfect measurement conditions
however. The question we focus on is what happens if we are only able
to refute it $97$ times and are uncertain $3$ times? This is quite
common in actual experiments. Mathematically we can model this idea
by stating that the probability of the event is in the range $\left[0,0.03\right]$
which says that the probability of the event could be $0$, $0.01$,
$0.02$, or $0.03$ as each the three missing records could either
be evidence for the event or against it. We just cannot nail it down
given the current experimental results and therefore use a ($\delta=$)$0.03$-deterministic
probability measure. The interesting observation is that the axioms
of probability theory (like additivity and convexity) impose enough
constraints on the structure of probability measures that make them
robust in the face of small $\delta$s.

@@

The situation changes sharply if the missing data reaches $\frac{1}{3}$,
which gives the following probability measure as a possible formalization:

\begin{equation}
\begin{aligned}\mu\left(\mathbb{0}\right) & =\imposs & \mu\left(\mathbb{1}\right) & =\necess\\
\mu\left(P\right) & =\left[0,\tfrac{1}{3}\right] & \mu\left(\mathbb{1}-P\right) & =\left[\tfrac{2}{3},1\right]
\end{aligned}
\,.
\end{equation}
This is not a valid probability measure for the same argument as above.
However, it might be the case that the missing data for one-dimensional
projectors always support the associated event while those for two-dimensional
projectors always refute the event. In this case, the same experimental
data is consistent with the following probability measure:

\begin{equation}
\begin{aligned}\mu\left(\mathbb{0}\right) & =\imposs & \mu\left(\mathbb{1}\right) & =\necess\\
\mu\left(P\right) & =\left[\tfrac{1}{3},\tfrac{1}{3}\right] & \mu\left(\mathbb{1}-P\right) & =\left[\tfrac{2}{3},\tfrac{2}{3}\right]
\end{aligned}
\,,
\end{equation}
which can be easily verified to be a valid QIVPM.

Similar argument can be applied on the same nine observables $\mathbf{O}_{ij}$
with $i$ and $j$ ranging over $\{0,1,2\}$ from the Mermin-Peres
``magic square'' used to prove the KS theorem \cite{Mermin1990Simple,peres1995quantum,Griffiths2003}:

@@

{\renewcommand{\arraystretch}{2}% 
\begin{center} 
\begin{tabular}{r|@{\quad}c@{\quad}|@{\quad}c@{\quad}|@{\quad}c@{\quad}|} 
$\mathbf{O}_{ij}$~ & $j=0$ & $j=1$ & $j=2$ \\ 
\hline  
$i=0~$ & $\mathbb{1}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\mathbb{1}$  & $\sigma_{z}\otimes\sigma_{z}$ \tabularnewline 
\hline  
$i=1~$ & $\sigma_{x}\otimes\mathbb{1}$  & $\mathbb{1}\otimes\sigma_{x}$  & $\sigma_{x}\otimes\sigma_{x}$ \tabularnewline 
\hline  
$i=2~$ & $\sigma_{x}\otimes\sigma_{z}$  & $\sigma_{z}\otimes\sigma_{x}$  & $\sigma_{y}\otimes\sigma_{y}$ \tabularnewline 
\hline  
\end{tabular}\,.
\par\end{center} 
}

\noindent The observables are constructed using the Pauli matrices
$\left\{ \mathbb{1},\sigma_{x},\sigma_{y},\sigma_{z}\right\} $ whose
eigenvalues are all either~$1$ or $-1$ \cite{Redhead1987-REDINA,544199,Griffiths2003,Jaeger2007,Mermin2007}.
They are arranged such that in each row and column, \emph{except the
column $j=2$}, every observable and its expectation value is the
product of the other two. In the $j=2$ column, we have instead that
$\left(\sigma_{z}\otimes\sigma_{z}\right)\left(\sigma_{x}\otimes\sigma_{x}\right)=-\sigma_{y}\otimes\sigma_{y}$
and 
\begin{equation}
\expval{\mathbf{O}_{02}}\expval{\mathbf{O}_{12}}=-\expval{\mathbf{O}_{22}}\,.\label{eq:j_2_expectation_value}
\end{equation}

Assume we measure each observable twice. By the KS theorem, it is
impossible to get the following measurement results because the product
of the expectation values in the $j=2$ column violate Eq.~(\ref{eq:j_2_expectation_value}).
\begin{center}
\begin{tabular}{cccc}
\toprule 
\addlinespace
$\mathbf{O}_{ij}$  & \multicolumn{2}{c}{Results} & $\expval{\mathbf{O}_{ij}}$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbf{O}_{00}$ , $\mathbf{O}_{01}$, $\mathbf{O}_{10}$ , $\mathbf{O}_{11}$  & $-1$ & $-1$ & $-1$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{02}$ , $\mathbf{O}_{12}$ , $\mathbf{O}_{20}$ , $\mathbf{O}_{21}$
, $\mathbf{O}_{22}$  & $1$ & $1$ & $1$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

Now consider what happens if if half of the records are missing, i.e.,
$\delta=\frac{1}{2}$, which gives the following measurement results,
where $\missing$ stands for missing data.
\begin{center}
\begin{tabular}{ccccc}
\toprule 
\addlinespace
$\mathbf{O}_{ij}$  & \multicolumn{2}{c}{Results} & $\begin{aligned} & \textrm{Missing}\\
 & \textrm{Record}\\
 & \textrm{Might Be}
\end{aligned}
$ & $\begin{aligned} & \textrm{Possible}\\
 & \expval{\mathbf{O}_{ij}}
\end{aligned}
$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbf{O}_{00}$, $\mathbf{O}_{11}$ & $-1$ & $\missing$ & $1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{01}$ , $\mathbf{O}_{10}$  & $\missing$ & $-1$ & $1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{02}$ , $\mathbf{O}_{20}$ , $\mathbf{O}_{22}$  & $1$ & $\missing$ & $-1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{12}$ , $\mathbf{O}_{21}$  & $\missing$ & $1$ & $-1$ & $0$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

\noindent Since the experimental results are missing, they might come
form the third column. In this kind of situation, the expectation
values of each observables is always $0$, and the product relations
among these expectation values must be consistent with the product
relations among observables.

### 2018-02-07 prop_prop_letter_ajhs_referee_response.tex

\date{\today}

\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In our QIVPM framework, the Kochen-Specker (KS) theorem is equivalent
to stating that there does not exist a ($\delta=$)$0$-deterministic
QIVPM. We further claim that there cannot exist any $\delta$-deterministic
QIVPM for $\delta<\frac{1}{3}$ and that there exist QIVPMs for every
$\delta\ge\frac{1}{3}$.

The referee questions the correctness of Thm.~\ref{cor:Kochen-Specker-IVPM}
and the physical interpretation of $\delta$-determinism. In answer
to the question of correctness, we attach the complete proof of Thm.~\ref{cor:Kochen-Specker-IVPM}
which we verified again in Appendix~\ref{sec:Proof-of-Theorem}.
In the next paragraphs, we explain the physical interpretation of
$\delta$-determinism and provide examples that illustrate why $\frac{1}{3}$
is special.

@@

We understand that the argument against $\delta$-determinism by the
referee assumes that the parameter $\delta$ is \emph{exclusively}
related to the density of points (states) defined in a general Hilbert
space. This is indeed Meyer's great insight and main assumption in
his proof about the field of rationals nullifying the KS theorem.
In other words, Meyer attributes finite-precision errors to the description
of the states. We, however, make no particular assumptions about the
description of the states, and attribute finite-precision errors to
the measurement process itself as realized in a laboratory. In our
framework, the parameter $\delta$ reflects insufficient knowledge
of the experimenter which could be due to a variety of reasons related
to imperfections of devices. A typical question that an experimenter
may not be able to answer accurately would be ``did an electron land
to the left or to the right of the screen''? There are cases in which
the electron would land too close to the middle of the screen for
the experimenter to be able to determine with certainty if that is
closer to the left or the right. That imprecision might be modeled
by limiting the description of the quantum states to rational points
but we do not make this assumption. We simply record this imprecision
in the probability we assign to events and use the axioms of probability
theory (additivity, etc) to argue that a sharp transition occurs when
the imprecisions reach a certain value.

The following argument shows a concrete but simplified situation which
explains the physical significance of the parameter $\delta$. We
then follow this explanation by providing experimental results for
a three-dimensional quantum system which cannot be described by any
$\delta$-deterministic QIVPM for $\delta<\frac{1}{3}$ but can be
described by a $\frac{1}{3}$-deterministic QIVPM.

The meaning of $\delta$: To determine the probability of any event,
we typically repeat an experiment $n$ times and count the number
of times we witness the event. This assumes that for each run of the
experiment we can determine, using our apparatuses, whether the event
occurred or not. Assume an event has an ideal mathematical probability
of $0$, and we repeat the experiment $100$ times. In a perfect world
we should be able to refute the event $100$ times and calculate that
the probability is $0$. We might also observe the event $2$ times
and refute it $98$ times and therefore calculate the probability
to be $0.02$. This situation still assumes perfect measurement conditions
however.

Let the quantum system be in state~$\ket{\psi}$ and the probability
of some event be $0$. Assume we repeat the experiment $100$ times
and we observe the event $1$ time. The Godsil-Zaks-Meyer approach
interprets this by assuming that the state was actually $\ket{\psi'}$
which is infinitesimally close to $\ket{\psi}$ and proceed to reason
about how coloring the states with rational points relates to the
coloring of arbitrary states. We make no such assumptions and do not
impose any constraints on the actual state of the system. In fact
we do not make any statements about such situations. What we are addressing
is a situation in which our devices do not allow to make a clear determination
of whether an event has happened or not. We model this by an uncertainty
in the probability of the event and then reason using the axioms of
probability theory to bound this uncertainty.

The question we focus on is what happens if we are only able to refute
it $97$ times and are uncertain $3$ times? This is quite common
in actual experiments. Mathematically we can model this idea by stating
that the probability of the event is in the range $\left[0,0.03\right]$
which says that the probability of the event could be $0$, $0.01$,
$0.02$, or $0.03$ as each the three missing records could either
be evidence for the event or against it. We just cannot nail it down
given the current experimental results and therefore use a ($\delta=$)$0.03$-deterministic
probability measure. The interesting observation is that the axioms
of probability theory (like additivity and convexity) impose enough
constraints on the structure of probability measures that make them
robust in the face of small $\delta$s.

@@

Now consider what happens if if half of the records are missing, i.e.,
$\delta=\frac{1}{2}$, which gives the following measurement results,
where $\missing$ stands for missing data.
\begin{center}
\begin{tabular}{ccccc}
\toprule 
\addlinespace
$\mathbf{O}_{ij}$  & \multicolumn{2}{c}{Results} & $\begin{aligned} & \textrm{Missing}\\
 & \textrm{Record}\\
 & \textrm{Might Be}
\end{aligned}
$ & $\begin{aligned} & \textrm{Possible}\\
 & \expval{\mathbf{O}_{ij}}
\end{aligned}
$\tabularnewline
\midrule
\midrule 
\addlinespace
$\mathbf{O}_{00}$, $\mathbf{O}_{11}$ & $-1$ & $\missing$ & $1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{01}$ , $\mathbf{O}_{10}$  & $\missing$ & $-1$ & $1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{02}$ , $\mathbf{O}_{20}$ , $\mathbf{O}_{22}$  & $1$ & $\missing$ & $-1$ & $0$\tabularnewline
\midrule 
\addlinespace
$\mathbf{O}_{12}$ , $\mathbf{O}_{21}$  & $\missing$ & $1$ & $-1$ & $0$\tabularnewline
\bottomrule
\end{tabular}
\par\end{center}

\noindent Since the experimental results are missing, they might come
form the third column. In this kind of situation, the expectation
values of each observables is always $0$, and the product relations
among these expectation values must be consistent with the product
relations among observables.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliography{prop}

\appendix

\section{\label{sec:Proof-of-Theorem}Proof of Theorem~\ref{cor:Kochen-Specker-IVPM}}

\begin{thm}[Finite-precision Extension of the Kochen-Specker Theorem]
\label{cor:Kochen-Specker-IVPM} Given a Hilbert space $\Hilb$ of
dimension~$d\ge3$, there is no $\delta$-deterministic QIVPM for
$\delta<\frac{1}{3}$.\end{thm}

The proof is by contradiction: Suppose there is a $\delta$-deterministic
QIVPM~$\bar{\mu}:\events\rightarrow\mathscr{I}$. Now use this assumed
QIVPM to construct the following $0$-deterministic QIVPM $\bar{\mu}^{\textrm{D}}:\events\rightarrow\left\{ \imposs,\necess\right\} $:
\begin{equation}
\bar{\mu}^{\textrm{D}}\left(P\right)=\begin{cases}
\imposs\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[0,\delta\right]\:;\\
\necess\,, & \textrm{ if }\bar{\mu}\left(P\right)\subseteq\left[1-\delta,1\right]\:.
\end{cases}
\end{equation}
However, since we know that $0$-deterministic QIVPMs do not exist,
the map $\bar{\mu}^{\textrm{D}}$ cannot exist and hence $\bar{\mu}$
does not exist. It remains to verify that $\bar{\mu}^{\textrm{D}}$
is indeed a QIVPM by checking that for orthogonal projectors $P_{0}$
and~$P_{1}$, 
\begin{equation}
\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\bar{\mu}^{\textrm{D}}\left(P_{0}\right)+\bar{\mu}^{\textrm{D}}\left(P_{1}\right)\,,\label{eq:QuantumInterval-valuedProbability-Equal}
\end{equation}
because Eq.~(\ref{eq:QuantumInterval-valuedProbability-Equal}) implies
the convexity condition.

When one of $\bar{\mu}^{\textrm{D}}\left(P_{0}\right)$ and $\bar{\mu}^{\textrm{D}}\left(P_{1}\right)$
is $\necess$, say $\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\necess$
and $\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$, we have $\bar{\mu}\left(P_{0}\right)\subseteq\left[1-\delta,1\right]$
and $\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$ which
implies $\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[1-\delta,1+\delta\right]$.
Since $\bar{\mu}\left(P_{0}+P_{1}\right)$ is a subset of $\left[0,1\right]$,
$\bar{\mu}\left(P_{0}+P_{1}\right)$ must be a subset of $\left[1-\delta,1\right]$,
which implies $\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\necess$.

When $\bar{\mu}^{\textrm{D}}\left(P_{0}\right)=\bar{\mu}^{\textrm{D}}\left(P_{1}\right)=\imposs$,
we have both $\bar{\mu}\left(P_{0}\right)$ and $\bar{\mu}\left(P_{1}\right)\subseteq\left[0,\delta\right]$
which implies $\bar{\mu}\left(P_{0}+P_{1}\right)\subseteq\left[0,2\delta\right]$.
Since $\bar{\mu}\left(P_{0}+P_{1}\right)$ is a subset of either $\left[0,\delta\right]$
or $\left[1-\delta,1\right]$ and the latter is excluded as illustrated
in Fig.~1, then $\bar{\mu}\left(P_{0}+P_{1}\right)$ must be a subset
of $\left[0,\delta\right]$, which implies $\bar{\mu}^{\textrm{D}}\left(P_{0}+P_{1}\right)=\imposs$.
\begin{widetext}
\begin{center}
\includegraphics[bb=50bp 100bp 900bp 300bp,clip,scale=0.5]{prop_prop_letter_ajhs_referee_response_pptx}\\
Figure 1.~\label{fig:Show-subset}Since $\bar{\mu}\left(P_{0}+P_{1}\right)$
and $\left[1-\delta,1\right]$ are disjoint, $\bar{\mu}\left(P_{0}+P_{1}\right)$
is a subset of $\left[0,\delta\right]$.
\par\end{center}
\end{widetext}
\end{document}

### 2018-02-07 prop_prop_letter_ajhs_referee_response.tex

\date{\today}

\maketitle
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

In our QIVPM framework, the Kochen-Specker (KS) theorem is equivalent
to stating that there does not exist a ($\delta=$)$0$-deterministic
QIVPM. We further claim that there cannot exist any $\delta$-deterministic
QIVPM for $\delta<\frac{1}{3}$ and that there exist QIVPMs for every
$\delta\ge\frac{1}{3}$.

The referee questions the correctness of Thm.~\ref{cor:Kochen-Specker-IVPM}
and the physical interpretation of $\delta$-determinism. In answer
to the question of correctness, we attach the complete proof of Thm.~\ref{cor:Kochen-Specker-IVPM}
which we verified again in Appendix~\ref{sec:Proof-of-Theorem} for
the referee's convenience. In the next paragraphs, we explain the
physical interpretation of $\delta$-determinism and provide examples
that illustrate why $\frac{1}{3}$ is special.

@@

The following argument shows a concrete but simplified situation which
explains the physical significance of the parameter $\delta$. We
then follow this explanation by providing experimental results for
a three-dimensional quantum system which cannot be described by any
$\delta$-deterministic QIVPM for $\delta<\frac{1}{3}$ but can be
described by a $\frac{1}{3}$-deterministic QIVPM.

The meaning of $\delta$: To determine the probability of any event,
we typically repeat an experiment $n$ times and count the number
of times we witness the event. This assumes that for each run of the
experiment we can determine, using our apparatuses, whether the event
occurred or not. Assume an event has an ideal mathematical probability
of $0$, and we repeat the experiment $100$ times. In a perfect world
we should be able to refute the event $100$ times and calculate that
the probability is $0$. We might also observe the event $2$ times
and refute it $98$ times and therefore calculate the probability
to be $0.02$. This situation still assumes perfect measurement conditions
however.

Let the quantum system be in state~$\ket{\psi}$ and the probability
of some event  be $0$. Assume we repeat the experiment $100$ times
and we observe the event $1$ time. The Godsil-Zaks-Meyer approach
interprets this by assuming that the state was actually $\ket{\psi'}$
which is infinitesimally close to $\ket{\psi}$ and proceed to reason
about how coloring the states with rational points relates to the
coloring of arbitrary states. We make no such assumptions and do not
impose any constraints on the actual state of the system. In fact
we do not make any statements about such situations. What we are addressing
is a situation in which our devices do not allow to make a clear determination
of whether an event has happened or not. We model this by an uncertainty
in the probability of the event and then reason using the axioms of
probability theory to bound this uncertainty. We should add that our
approach is agnostic about attributing elements of reality to the
state function (as Meyer). We simply attribute lack of certainty to
the measured results through the use of an instrument.
